---
title: "Tornado hit people, thunderstorms hurt economics"
author: "Alex Gaggin"
date: "Friday, August 14, 2015"
output: html_document
---

##Synopsis

U.S. National Oceanic and Atmospheric Administration (NOAA)
tracks and classifies significant weather events in the United States in their
storm database,
assessing, in particular, human and dollar damages. The data
in the dataset was available for years 1950-2011, but for consistency reasons
a shorter time period was analyzed - 1996-2010. The conclusion appears valid for
longer period as well: tornadoes were the threat to human life and health even
before unusually deadly outbreak of 2011.
From economic damages point of view, tornadoes were not as bad
as thunderstorms and flash floods.

## Data Processing

```{r, message=FALSE, warning=FALSE}
# Preload libraries
library(digest)
library(dplyr)
library(lubridate)
library(ggplot2)
```

The data was prepared for download by authors of Reproducible Research course on
the Coursera. It's downloaded and loaded to an R data frame.

```{r, cache=TRUE}
# setwd(file.path(normalizePath("~"),"Coursera-R","RepData_PeerAssessment2"))
# Knitr works in the Rmd file directory, so this isn't needed for it, but
# it can be useful to re-initialize session in console for dev/test/debug

# Download the file - ignore "%2F" imperfection when making a local file name
remote<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
localzip<-basename(remote)
if(!file.exists(localzip)) download.file(remote,localzip)
# No need to unzip - read.csv() does it automatically

# For chunk's console usage - no need to reload data if it's loaded and intact
if(!exists("loaded") || digest(loaded) != "b223723dffbf2ec857526bcc32d2a31d")
   loaded <- read.csv(localzip)
# We keep original data frame for console work caching and make a mutable copy
s<-loaded
```

Beginning dates of the events are converted to the time class and year part
is extracted to a separate variable.

```{r, cache=TRUE}
# Convert beginning dates to date format
s$BGN_DATE<-mdy_hms(s$BGN_DATE)
s$Year<-format(s$BGN_DATE,"%Y")
```

We will consider not just fatalities and property damage, but health damage
and economic damage overall, so we need to add different types of damages
together.

```{r, cache=TRUE}
s$HealthDamage<-s$FATALITIES+s$INJURIES
s$EconomicDamage<-s$PROPDMG+s$CROPDMG
```

Event type names seem to duplicate, let's aggregate them.

```{r}
# After checking unique(s$EVTYPE)
s$type<-tolower(s$EVTYPE)
aggr<-function(df,pattern,to)
        df %>% mutate(type=ifelse(grepl(pattern,type,fixed=FALSE),to,type))
s<-aggr(s,"tstm","thunderstorm")
s<-aggr(s,"thunderstorm","thunderstorm")

s<-aggr(s,"wind","wind")
s<-aggr(s,"wnd","wind")

s<-aggr(s,"slide","debris flow")
s<-aggr(s,"lands","debris flow")

s<-aggr(s,"nado","tornado")
s<-aggr(s,"ndao","tornado")

s<-aggr(s,"flash fl","flash")

s<-aggr(s,"flood","flood")
s<-aggr(s,"fld","flood")
s<-aggr(s,"high water","flood")

s<-aggr(s,"hail","hail")

s<-aggr(s,"hail","hail")

s<-aggr(s,"spout","spouts")

s<-aggr(s,"cold","cold")
s<-aggr(s,"frost","cold")
s<-aggr(s,"freez","cold")
s<-aggr(s,"low t","cold")
s<-aggr(s,"hypoth","cold")

s<-aggr(s,"snow","snow")

s<-aggr(s,"lightn","lightning")

s<-aggr(s,"blizz","blizzard")

s<-aggr(s,"rain","rain")
s<-aggr(s,"shower","rain")
s<-aggr(s,"prec","rain")
s<-aggr(s,"wet","rain")

s<-aggr(s,"warm","heat")
s<-aggr(s,"heat","heat")
s<-aggr(s,"hot","heat")
s<-aggr(s,"record high","heat")

s<-aggr(s,"dry","drought")
s<-aggr(s,"drie","drought")
s<-aggr(s,"droug","drought")

s<-aggr(s,"tide","tide")
s<-aggr(s,"surf","tide")
s<-aggr(s,"wave","tide")
s<-aggr(s,"coastal surge","tide")
s<-aggr(s,"high seas","tide")

s<-aggr(s,"volcan","volcano")

s<-aggr(s,"ice","ice")
s<-aggr(s,"icy","ice")

s<-aggr(s,"winter","winter")
s<-aggr(s,"wintry","winter")

s<-aggr(s,"fire","fires")

s<-aggr(s,"swell","swells")

s<-aggr(s,"hurr","hurricane")
s<-aggr(s,"typh","hurricane")
s<-aggr(s,"tropical","hurricane")

s<-aggr(s,"dust","dust")

s<-aggr(s,"temper","temperature")

s<-aggr(s,"flash","flash flood")
```

The NOAA Storm Events Database
[webpage](https://www.ncdc.noaa.gov/stormevents/details.jsp) explains that
before 1955 the only event type registered was tornado, and then until 1996
three types of events were recorded - tornado, thunderstorm wind and hail.
Thus the comparison for event types should start in 1996.

```{r,cache=TRUE}
recent<-s[s$BGN_DATE>ymd("1996-01-01"),]
```

What are top 10 event types that caused most life loss and injuries?

```{r}
deadly10<-tapply(recent$HealthDamage,recent$type,sum) %>%
        sort(decr=TRUE) %>%
        head(10) %>%
        data.frame
names(deadly10)<-'HealthDamage'
deadly10$Event<-rownames(deadly10)
# Order it to prevent gplot from re-sorting bar alphabetically
deadly10$Event<-factor(deadly10$Event, levels=rev(deadly10$Event))
print(deadly10, row.names=FALSE)
```


What are top event types that caused most economic damage?

```{r}
costly10<-tapply(recent$EconomicDamage,recent$type,sum) %>%
        sort(decr=TRUE) %>%
        head(10) %>%
        data.frame
names(costly10)<-'EconomicDamage'
costly10$Event<-rownames(costly10)
costly10$Event<-factor(costly10$Event, levels=rev(costly10$Event))
print(costly10, row.names=FALSE)
```


Aggregate health damage cost by tornado annually.

```{r}
Tornado<-recent[recent$type=="tornado",]
Annual.tornado.health.damage<-with(Tornado, tapply(HealthDamage,Year,sum))
# Compendate for the missing month in 2011
Annual.tornado.health.damage["2011"]<-
        round(Annual.tornado.health.damage["2011"]*12/11)
Annual.tornado.health.damage
```

There's an outlier in 2011 (a quick google search shows there was indeed
a deadliest tornado outbreak in April 2011), let's see damages
in 1996 - 2010 instead to set the outbreak aside.

```{r,cache=TRUE}
framed<-recent[recent$BGN_DATE<ymd("2011-01-01"),]

deadly10<-tapply(framed$HealthDamage,framed$type,sum) %>%
        sort(decr=TRUE) %>%
        head(10) %>%
        data.frame
names(deadly10)<-'HealthDamage'
deadly10$Event<-rownames(deadly10)
# Order it to prevent gplot from re-sorting bar alphabetically
deadly10$Event<-factor(deadly10$Event, levels=rev(deadly10$Event))
print(deadly10, row.names=FALSE)

costly10<-tapply(framed$EconomicDamage,framed$type,sum) %>%
        sort(decr=TRUE) %>%
        head(10) %>%
        data.frame
names(costly10)<-'EconomicDamage'
costly10$Event<-rownames(costly10)
costly10$Event<-factor(costly10$Event, levels=rev(costly10$Event))
print(costly10, row.names=FALSE)
```

It looks like we can operate 1996-2010 year range to draw general conclusions,
as dropping 2011 doesn't change the event type comparison picture. Before
1996 the data was focused on tornadoes with thunderstorms added later, so
we shouldn't expect any possible changes to our conclusions from that data.

While it's unreasonable to expect for life loss to be measured in any units
other than number of people, for economic damage the scale can be anything.
Let's figure out what scale the economic damages is in our dataset. To do this,
let's pick an event and find it in an unprocessed file from NOAA's website.

```{r, cache=TRUE}
# Pick up some recent event with economic damage and begin latitude
somedmg<-tail(framed[framed$EconomicDamage>0 & framed$LATITUDE != 0 ,],1)

# Look at its year, latitude, location and property damage
somedmg %>% select(Year,LATITUDE,BGN_LOCATI,PROPDMG)

# Get unprocessed file with event details for 2010, read it, look at it
url<-"http://www1.ncdc.noaa.gov/pub/data/swdi/stormevents/csvfiles/StormEvents_details-ftp_v1.0_d2010_c20140824.csv.gz"
d2010<-basename(url)
if (!file.exists(d2010)) download.file(url,d2010)
bit<-read.csv(d2010)
names(bit)

# Try to match by latitude
sum(bit$BEGIN_LAT %in% somedmg$LATITUDE)

# Nothing? Try by location
sum(bit$BEGIN_LOCATION %in% somedmg$BGN_LOCATI)

# There's only one event, let's check it out
origdmg<-bit[bit$BEGIN_LOCATION %in% somedmg$BGN_LOCATI,]

origdmg %>% select(-EPISODE_NARRATIVE)

somedmg %>% select(-REMARKS)

# Seems to be the same event, not sure why coordinates are a bit different -
# was there an update to the database since the copy for the task was made
# by the author of the pre-processed dataset?

# What's the scale of damages?
origdmg$DAMAGE_PROPERTY
```

Thus in our dataset property damage is measured in thousands of dollars.

## Results

Tornadoes, supposedly as a result of their rapid nature, are the biggest
cause of human life loss and injuries. They are far ahead even without
significant outbreak in 2011.

```{r}
redplot<-ggplot(data=deadly10,
          aes(x=Event,y=HealthDamage)
          )+
        geom_bar(stat='identity',fill='red')+
        coord_flip()+
        ggtitle("Top 10 types of weather events that caused\n biggest loss of human life and health in 1996-2010")+
        ylab("Number of people killed or injured")+
        xlab("Event type")
print(redplot)
```


Damage to property and agriculture is more significant from thunderstorms and
flash floods - these aren't as deadly, but supposedly hit much larger areas.

```{r}
# Scale damages to thousands to billions
greenplot<-ggplot(data=costly10,
          aes(x=Event,y=EconomicDamage/1000000)
          )+
        geom_bar(stat='identity',fill='darkgreen')+
        coord_flip()+
        ggtitle("Top 10 types of weather events that caused\n most damage to property and crops in 1996-2010")+
        ylab("Economic damage, $US billion")+
        xlab("Event type")
print(greenplot)

```

<!--
## To do
- Check and handle NAs
- Do a geo check
- Should I treat outliers? Do they and regular events need different policies?
-->